## Identifying corridors

### Overview

Overview of the importance of identifying and mapping seasonal movement corridors for caribou.

### Walkthrough

In this section we describe the steps used to estimate individual home ranges using the KDE approach and compare it to another commonly used method, the minimum convex hull (MCP). We illustrate the approach for one individual caribou.

```{r}
#| echo: false
#| warning: false
#| message: false
library(sf)
library(amt)
library(tmap)
library(terra)
library(tidyverse)
```

#### Step 1. Identify migration periods

This step is described in "Seasonal migrations" in the "Defining segments" section.

#### Step 2. Map individual paths

Simply connecting the dots of a migration sequence with a straight line gives a rough idea of the route, but it's an oversimplification. It doesn't reveal the true width of the path or, more importantly, the specific areas where an animal stopped-over to rest and feed. To build a more realistic map, we used a statistical method called the Brownian Bridge Movement Model (BBMM). This model is incorporates a key parameter called the Brownian motion variance (BMV), which is a statistical estimate of an individual's mobility, making the resulting map more realistic. The output of this model is a utilization distribution (UD), which is essentially a "heat map" or "probability surface." Instead of just showing a line, the UD shows the likelihood of an individual having been in any given location along its route. The "hotter" the color, the more time the individual likely spent there.

Using a BBMM to create a utilization distribution has two major benefits:

- Estimates path width: The model calculates the probable width of the movement path around the straight lines connecting the points. This transforms a simple line into a more realistic two-dimensional corridor, showing the full area the individual may have used.
- Identifies stopover sites: The "hottest" parts of the heat map are incredibly valuable. They reveal stopover sites - critical areas where animals spend extended time foraging and resting during their long journey. These sites are as important to the success of a migration as the path itself.

```{r}
#| echo: false
#| warning: false
#| message: false
#| fig-width: 9
#| fig-height: 6
#| fig-cap: "Figure 1. Brownian bridge movement model (95% contour) for one caribou in 2021."
gps <- read_csv("../data/little_rancheria/caribou.csv")
aoi <- st_read("../data/little_rancheria/corridors.gpkg", "aoi", quiet=TRUE)

# Spring
gps1s <- gps |> filter(id=="SKRATALR0018" & year==2021 & migration=="Spring migration")
gps1s_sf <- gps1s |>
  st_as_sf(coords=c("longitude", "latitude"), crs=4326, remove=FALSE)
xs <- st_read("../data/little_rancheria/corridors.gpkg", "bbmm95", quiet=TRUE) |>
  filter(id=="43141" & year==2021 & migration=="Spring migration")

# Fall
gps1f <- gps |> filter(id=="SKRATALR0018" & year==2021 & migration=="Fall migration")
gps1f_sf <- gps1f |>
  st_as_sf(coords=c("longitude", "latitude"), crs=4326, remove=FALSE)
xf <- st_read("../data/little_rancheria/corridors.gpkg", "bbmm95", quiet=TRUE) |>
  filter(id=="43141" & year==2021 & migration=="Fall migration")

ms <- tm_basemap("Esri.WorldTopoMap") +
  tm_shape(aoi) +
  tm_borders() +
  tm_shape(xs) +
  tm_polygons(fill.legend = tm_legend_hide()) +
  tm_shape(gps1s_sf) +
  tm_dots(fill="red") +
  tm_title("Spring migration")
mf <- tm_basemap("Esri.WorldTopoMap") +
  tm_shape(aoi) +
  tm_borders() +
  tm_shape(xf) +
  tm_polygons(fill.legend = tm_legend_hide()) +
  tm_shape(gps1f_sf) +
  tm_dots(fill="red") +
  tm_title("Fall migration")
tmap_arrange(ms, mf)
```

#### Step 3. Average individual paths

To protect a migration, we need to see the main corridor used by the entire population, not just the path of one individual. This can be achieved this by combining all the individual maps into a single, comprehensive view of the herd's movement. This involves a three step process:

- Average individual migrations: First, for each individual, we took all of its yearly migration UDs (the heat maps) and average them together. This creates a single map representing the typical migration path for that specific animal over time.
- Define individual footprints: For each animal's averaged map, we defined its "footprint" as the area that contains 99% of its probable locations. This captures the vast majority of where that individual traveled.
- Stack the footprints: All of these individual footprints are then layered, or "stacked" on top of one another. The result is a map where every pixel on the grid has a value corresponding to how many different individuals passed through it.

```{r}
#| echo: false
#| warning: false
#| message: false
#| fig-width: 9
#| fig-height: 6
#| fig-cap: "Figure 2. Movement paths based on BBMM models for all caribou in all years."
rs <- rast("../data/little_rancheria/spring_migration.tif")
ms <- tm_basemap("Esri.WorldTopoMap") +
  tm_shape(aoi) +
  tm_borders() +
  tm_shape(rs) +
  tm_raster(col.legend = tm_legend(
    title = "Caribou use", position = tm_pos_in(pos.v = "bottom", pos.h="right"))) +
  tm_title("Spring migration")
rf <- rast("../data/little_rancheria/fall_migration.tif")
mf <- tm_basemap("Esri.WorldTopoMap") +
  tm_shape(aoi) +
  tm_borders() +
  tm_shape(rf) +
  tm_raster(col.legend = tm_legend(
    title = "Caribou use", position = tm_pos_in(pos.v = "bottom", pos.h="right"))) +
  tm_title("Fall migration")
tmap_arrange(ms, mf)
```

#### Step 4. Classify corridors

Finally, we can classify the resultant stacked map to visualize how 1) intensely different parts of the corridor are used by the herd and 2) where the key stopovers areas are located. This allows us to distinguish between occasional and essential parts of the corridor.

**Output 1: Corridor use intensity**

The levels of corridor use are typically defined as follows:

| Use Level      | Definition                                      | Importance                                                                    |
| ---------- | ------------------------------------------------| ------------------------------------------------------------------------- |
| Low Use    | Areas used by at least one individual. | Shows the maximum extent of the entire migration footprint.                   |
| Medium Use | Areas used by 10-20% of the individuals.        | Represents regularly used routes.                                             |
| High Use   | Areas used by >20% of individuals.              | Pinpoints the most critical, core pathways essential for the herd's survival. |

<br>
**Output 2: High use areas**

High use areas or stopovers are areas where animals spend extended time foraging and resting during migration. They are defined as the top 10% of use values from the propulation-level probability map. Protecting these high-use areas is as important as protecting the connectivity between them.

```{r}
#| echo: false
#| warning: false
#| message: false
#| fig-width: 9
#| fig-height: 6
#| fig-cap: "Figure 3. High use areas within movement paths."
rs <- rast("../data/little_rancheria/spring_highuse.tif")
ms <- tm_basemap("Esri.WorldTopoMap") +
  tm_shape(aoi) +
  tm_borders() +
  tm_shape(rs) +
  tm_raster(col.legend = tm_legend(
    title = "High use", position = tm_pos_in(pos.v = "bottom", pos.h="right"))) +
  tm_title("Spring migration")
rf <- rast("../data/little_rancheria/fall_highuse.tif")
mf <- tm_basemap("Esri.WorldTopoMap") +
  tm_shape(aoi) +
  tm_borders() +
  tm_shape(rf) +
  tm_raster(col.legend = tm_legend(
    title = "High use", position = tm_pos_in(pos.v = "bottom", pos.h="right"))) +
  tm_title("Fall migration")
tmap_arrange(ms, mf)
```