## Traditional Habitat Selection

### Introduction

Fitting models to estimate traditional habitat selection functions (HSF) at the third-order of selection (e.g. selection within the home range).


```{r}
#| echo: FALSE
#| message: FALSE
#| warning: FALSE
library(glmmTMB)
library(sf)
library(amt)
library(terra)
library(tmap)
library(tidyverse)
```

### Prepare data

Note: This section can be moved to its own file eventually

**Movement data**

```{r}
#| echo: FALSE
#| message: FALSE
#| warning: FALSE
gps <- read_csv("../../data/little_rancheria/caribou.csv") |>
  filter(migration=="Fall migration")
trk <- gps |> make_track(.x=longitude, .y=latitude, .t=timestamp, id=id, year=year, crs=4326) |>
  transform_coords("EPSG:3578")
xbnd <- hr_kde(trk, levels=0.999) |>
  hr_isopleths()
x <- st_read("../../data/little_rancheria/caribou.gpkg", "gps", quiet=TRUE) |>
  filter(migration=="Fall migration") |>
  select(id) |>
  st_transform(3578)
#y <- x |> sample_n(500)
tm_shape(xbnd) + tm_borders() + tm_shape(x) + tm_dots()
```

**Prepare covariates**

```{r}
#| echo: FALSE
#| message: FALSE
#| warning: FALSE

# Forest proportion
lcc <- rast("../../data/little_rancheria/nalc2020.tif") |>
  crop(xbnd, mask=TRUE)
forest <- ifel(lcc<4, 1, 0)
for500 <- focal(forest, w=29, fun="sum")

# Distance to footprint
fp <- st_read("../../data/little_rancheria/disturbances.gpkg", "footprint_500m", quiet=TRUE)
fp.dist <- rasterize(fp, lcc) |>
  distance() |>
  crop(xbnd, mask=TRUE)

# Distance to burn
burn <- rast("../../data/little_rancheria/burns.tif") |>
  resample(lcc)
burn.dist <- burn |> distance() |>
  crop(xbnd, mask=TRUE)

r <- c(for500, fp.dist, burn.dist) |>
#r <- c(forest, fp.dist, burn.dist) |>
  scale()
names(r) <- c("for500", "fp.dist", "burn.dist")
plot(r)
```

**Extract covariates**

```{r}
#| echo: FALSE
#| message: FALSE
#| warning: FALSE

# Used locations
x1 <- terra::extract(r, x, bind=TRUE, ID=FALSE) |>
  as_tibble() |> 
  mutate(status=1)

# Available locations
ids <- unique(x$id)
wts <- as.numeric(table(x$id))
x0 <- spatSample(r, nrow(x)*10, na.rm=TRUE) |>
  mutate(id=sample(ids, nrow(x)*10, replace=TRUE, prob=wts), 
         status=0) |>
  as_tibble()
x10 <- rbind(x1, x0)
x10
```

### First model

Let’s fit a model to one individual’s used and available data

```{r}
#| echo: FALSE
#| message: FALSE
#| warning: FALSE
i18 <- x10 |>
  filter(id=="SKRATALR0018")
m1 = glmmTMB(status ~ for500 + fp.dist + burn.dist, family=binomial(), data=i18)
summary(m1)
```
  
Let's first consider the "Intercept". This parameter is largely the ratio of used to available sample. If we increase our available sample, this estimate will get smaller. Its value has a lot to do with how we setup the data to approximate the true underlying model. It has no biological interpretation, so we can skip it.

Next, we have estimated the effect of "for500" as -0.68. This estimate is negative, indicating that as the value of "for500" increases from its mean (defined at 0) while the values of `forest` and `shrub` (the other variables in the model) are held at their mean that habitat selection decreases. In other words, habitat use relative to what is available to this individual (as we defined it!) decreases the greater the proportion of forest. That's a lot of qualifiers to understand this estimate. These are important though. In terms of evidence of an effect, we can say that at a defined probability of Type I error of 0.05, we have a [statistically clear](https://doi.org/10.1111/2041-210X.13159) effect and that that it is not likely zero. The evidence of this is the very small p-value of ?.

Next, we have estimated the effect of `forest` as 0.12. This estimate is positive, indicating that as the value of `forest` increases from its mean (defined at 0) while the values of `dist.dev` and `shrub` (the other variables in the model) are held at their mean that habitat selection increases relative to what is available to this individual. At our defined Type I error we do not have statistical clarity and there is a reasonable chance that it is zero. (p-value = 0.4665834).

Next, we have estimated the effect of `shrub` as 0.88. This estimate is positive, indicating that as the value of `shrub` increases from its mean (defined at 0) while the values of `dist.dev` and `forest` (the other variables in the model) are held at their mean that habitat selection increases relative to what is available to this individual. At our defined Type I error we do have statistical clarity of an effect that is not zero (p-value = 8.2627753\times 10\^{-7}).

**Relative Selection**

How do we quantitatively evaluate two locations in terms of habitat selection using our model results? We can do so using Equation 8 of Fieberg et al. 2021.

Perhaps we want to compare a location that is very near development (dist.dev = -2) at high forest and shrub cover (forest = 2, shrub = 2) with that of a location far from development (dist.dev = 2) and also at high forest but low shrub cover (forest =2, shrub = -2).

```{r}
#| echo: FALSE
#| message: FALSE
#| warning: FALSE
# Get estimates without the intercept
coef = fixef(m1)[[1]][-1]

RS = exp(-2*coef[1] + 2*coef[2] + 2*coef[3]) / 
     exp(2*coef[1] + 2*coef[2] + -2*coef[3])
RS
```

If given equivalent availability between these two types of sites ( 1) near development at high forest and shrub cover, 2) far from development at high forest and shrub cover ), this individual would relatively select the first location by a factor of 2801.05.

Why are we exponentiating our linear combinations of terms (additions of covariates times estimated coefficients)? This is because in this setup, we have assumed the habitat selection process follows an exponential form (see section 10. The model being approximated) and equation 1 of Gerber and Northrup, 2020.

Relative Selection Strength

Avgar et al. 2017 refers to the relative selection strength (RSS) as the exponentiation of each coefficient.

For example,

```{r}
#| echo: FALSE
#| message: FALSE
#| warning: FALSE
# Coefficient for dist.dev
exp(coef[2])
```

Given two locations that differ by 1 standard deviation of dist.dev, but are otherwise equal, this individual would be 1.1324727 as likely to choose the location with higher dist.dev (or, equivalently, 0.8830235 times more likely to choose the location with the lower dist.dev.

#### Prediction

Lets combine our covariates now to predict relative habitat selection over more scenarios. To do this, we need to remember we used generalized linear modeling functions to approximate our model. We need to take care when predicting. First, we need to remember to drop the intercept. Second, we need to remember our true link function for relative habitat selection is the log-link (inverse-link is the exponential function) and not the logit-link, which is the default when fitting logistic regression models.

```{r}
#| echo: false
#| warning: false
#| message: false
# Force the intercept to be zero without any error
m1$fit$par[1] = 0
m1$sdr$par.fixed[1] = 0
m1$sdr$cov.fixed[1,] = 0
m1$sdr$cov.fixed[,1] = 0

# Create the dataframe for the combinations of variables for prediction  
newdata=data.frame(for500 = mean(i18$for500),
                   fp.dist = seq(-2,2, by=0.1),
                   #fp.dist = seq(min(i18$fp.dist), max(i18$fp.dist), by=0.1),
                   burn.dist = mean(i18$burn.dist)
                   )
# Next, predict 
preds = predict(m1,
                type = "link", # NOTE: this is where we tell the 'predict' function to use the same link function as the habitat selection function we fit above
                newdata = newdata, 
                se.fit = TRUE
                )
# Use the Normal Approximation to get confidence intervals (other approachces are possible too!)  
preds$LCL = preds$fit-1.96*preds$se.fit
preds$UCL = preds$fit+1.96*preds$se.fit

preds = data.frame(preds)  

# Exponentiate predicted values
preds.exp = exp(preds[,-2])  
```

**fp.dist**

```{r}
#| echo: false
#| warning: false
#| message: false
# Plot
plot(newdata$fp.dist,preds.exp$fit,type="l",lwd=3,xlab="Distance to Development (Scaled)",
     ylab="Relative Intensity of Selection")
lines(newdata$fp.dist,preds.exp$LCL,lwd=3,col=2,lty=4)
lines(newdata$fp.dist,preds.exp$UCL,lwd=3,col=2,lty=4)
abline(h=1, lwd=4, col=4)
text(1.5,1.4, "Selection")
text(1.5,0.6, "Avoidance")
```

We see that for this individual, given equal availability and `forest` and `shrub` at their mean values, avoid areas far from development (high values on x-axis) and select for areas close to development (low values on x-axis).

We can use this same process to extract values across spatial layers to predict relative habitat selection that is mapped.

**burn.dist**

```{r}
#| echo: false
#| warning: false
#| message: false
# Force the intercept to be zero without any error
m1$fit$par[1] = 0
m1$sdr$par.fixed[1] = 0
m1$sdr$cov.fixed[1,] = 0
m1$sdr$cov.fixed[,1] = 0

# Create the dataframe for the combinations of variables for prediction  
newdata=data.frame(for500 = mean(i18$for500),
                   fp.dist = mean(i18$fp.dist),
                   #fp.dist = seq(min(i18$fp.dist), max(i18$fp.dist), by=0.1),
                   burn.dist = seq(-2,2, by=0.1))
                   
# Next, predict 
preds = predict(m1,
                type = "link", # NOTE: this is where we tell the 'predict' function to use the same link function as the habitat selection function we fit above
                newdata = newdata, 
                se.fit = TRUE
                )
# Use the Normal Approximation to get confidence intervals (other approachces are possible too!)  
preds$LCL = preds$fit-1.96*preds$se.fit
preds$UCL = preds$fit+1.96*preds$se.fit

preds = data.frame(preds)  

# Exponentiate predicted values
preds.exp = exp(preds[,-2])  

# Plot
plot(newdata$burn.dist,preds.exp$fit,type="l",lwd=3,xlab="Distance to burned areas (Scaled)",
     ylab="Relative Intensity of Selection")
lines(newdata$burn.dist,preds.exp$LCL,lwd=3,col=2,lty=4)
lines(newdata$burn.dist,preds.exp$UCL,lwd=3,col=2,lty=4)
abline(h=1, lwd=4, col=4)
text(1.5,1.4, "Selection")
text(1.5,0.6, "Avoidance")
```

**for500**

```{r}
#| echo: false
#| warning: false
#| message: false
# Force the intercept to be zero without any error
m1$fit$par[1] = 0
m1$sdr$par.fixed[1] = 0
m1$sdr$cov.fixed[1,] = 0
m1$sdr$cov.fixed[,1] = 0

# Create the dataframe for the combinations of variables for prediction  
newdata=data.frame(for500 = seq(-2,2, by=0.1),
                   fp.dist = mean(i18$fp.dist),
                   #fp.dist = seq(min(i18$fp.dist), max(i18$fp.dist), by=0.1),
                   burn.dist = mean(i18$for500))
                   
# Next, predict 
preds = predict(m1,
                type = "link", # NOTE: this is where we tell the 'predict' function to use the same link function as the habitat selection function we fit above
                newdata = newdata, 
                se.fit = TRUE
                )
# Use the Normal Approximation to get confidence intervals (other approachces are possible too!)  
preds$LCL = preds$fit-1.96*preds$se.fit
preds$UCL = preds$fit+1.96*preds$se.fit

preds = data.frame(preds)  

# Exponentiate predicted values
preds.exp = exp(preds[,-2])  

# Plot
plot(newdata$for500,preds.exp$fit,type="l",lwd=3,xlab="Proportion forested (Scaled)",
     ylab="Relative Intensity of Selection")
lines(newdata$for500,preds.exp$LCL,lwd=3,col=2,lty=4)
lines(newdata$for500,preds.exp$UCL,lwd=3,col=2,lty=4)
abline(h=1, lwd=4, col=4)
text(1.5,1.4, "Selection")
text(1.5,0.6, "Avoidance")
```
