---
title: Connectivity Mapping Workflow Using amt
---

## Objective

Estimate and map connectivity between seasonal ranges for caribou using:

- GPS telemetry,
- integrated step-selection functions (iSSF),
- landcover and burn-history covariates,
- and resistance-based corridor modeling.

## 1. Load packages

```{r}
#| echo: FALSE
#| message: FALSE
#| warning: FALSE
library(amt)
library(sf)
library(terra)
library(dplyr)
library(lubridate)
library(survival)
library(gdistance)
library(tmap)
```

## 2. Load telemetry data

Assume columns:

- id
- timestamp
- longitude
- latitude

```{r}
#| echo: FALSE
#| message: FALSE
#| warning: FALSE
tele <- read.csv("caribou_fall_migration.csv")

tele$timestamp <- ymd_hms(tele$timestamp)

# Convert to sf
tele_sf <- st_as_sf(
  tele,
  coords = c("longitude", "latitude"),
  crs = 4326
)

# Project to UTM
utm_crs <- 3578   # example only — change if needed

tele_sf <- st_transform(tele_sf, crs = utm_crs)

coords <- st_coordinates(tele_sf)

tele$X <- coords[,1]
tele$Y <- coords[,2]

trk <- tele |> make_track(.x=longitude, .y=latitude, .t=timestamp, id=id, year=year, crs=4326) |>
  transform_coords("EPSG:3578")
xbnd <- hr_kde(trk, levels=0.999) |>
  hr_isopleths()

```


## 3. Load covariate rasters

```{r}
#| echo: FALSE
#| message: FALSE
#| warning: FALSE
# Forest proportion
lcc <- rast("../../data/little_rancheria/nalc2020.tif") |>
  crop(xbnd, mask=TRUE)
forest <- ifel(lcc<4, 1, 0)
for500 <- focal(forest, w=29, fun="sum")

# Distance to footprint
fp <- st_read("../../data/little_rancheria/disturbances.gpkg", "footprint_500m", quiet=TRUE)
fp.dist <- rasterize(fp, lcc) |>
  distance() |>
  crop(xbnd, mask=TRUE)

# Distance to burn
burn <- rast("../../data/little_rancheria/burns.tif") |>
  resample(lcc)
burn.dist <- burn |> distance() |>
  crop(xbnd, mask=TRUE)

# Stack rasters
covs <- c(for500, fp.dist, burn.dist) # |> scale()
names(covs) <- c("for500", "fp.dist", "burn.dist")
```

## 4. Create movement track

```{r}
#| echo: FALSE
#| message: FALSE
#| warning: FALSE
trk <- make_track(
  tele,
  .x = X,
  .y = Y,
  .t = timestamp,
  id = id,
  crs = utm_crs
)
```

## 5. Resample telemetry

Choose interval appropriate for migration behavior.

Example: 6-hour fixes.

```{r}
#| echo: FALSE
#| message: FALSE
#| warning: FALSE
trk_rs <- trk %>%
  track_resample(
    rate = hours(6),
    tolerance = minutes(30)
  ) %>%
  filter_min_n_burst()
```

## 6. Generate observed + random steps

```{r}
#| echo: FALSE
#| message: FALSE
#| warning: FALSE
steps <- steps_by_burst(trk_rs)

# Generate 20 available steps per used step
rand_steps <- random_steps(
  steps,
  n = 20
)
```

## 7. Extract covariates

```{r}
#| echo: FALSE
#| message: FALSE
#| warning: FALSE
steps_cov <- extract_covariates(
  rand_steps,
  covs
)
```

## 8. Prepare covariates

Convert landcover to factor if necessary.

```{r}
#| echo: FALSE
#| message: FALSE
#| warning: FALSE
steps_cov <- steps_cov %>%
  mutate(
    #landcover = factor(landcover),
    log_sl = log(sl_),
    cos_ta = cos(ta_)
  )
```

## 9. Fit integrated Step Selection Function (iSSF)

Example model:

movement kernel:
- step length
- turning angle

habitat selection:
- forest cover proportion
- distance to fooprint
- distance to burned area

```{r}
#| echo: FALSE
#| message: FALSE
#| warning: FALSE
# "for500", "fp.dist", "burn.dist"
issf_fit <- fit_issf(
  case_ ~
    for500 +
    fp.dist +
    burn.dist +
    log_sl +
    cos_ta +
    strata(step_id_),
  data = steps_cov
)

summary(issf_fit)
```

## 10. Interpret coefficients

Typical interpretation:

- positive burn coefficient:
  - selection for burned areas
- negative burn coefficient:
  - avoidance of burns
- landcover coefficients:
  - relative selection among vegetation types

## 11. Build habitat suitability / movement surface

Predict relative movement probability across landscape.

Extract raster values

```{r}
#| echo: FALSE
#| message: FALSE
#| warning: FALSE
#cov_df <- as.data.frame(covs, xy = TRUE, na.rm = FALSE)
cov_df <- as.data.frame(covs, xy = TRUE, na.rm = TRUE)
#cov_df$landcover <- factor(cov_df$landcover)
```

## 12. Predict linear predictor

```{r}
#| echo: FALSE
#| message: FALSE
#| warning: FALSE
cov_df$log_sl <- mean(steps_cov$log_sl, na.rm = TRUE)
cov_df$cos_ta <- mean(steps_cov$cos_ta, na.rm = TRUE)

beta <- coef(issf_fit)
cov_df$pred <- as.numeric(as.matrix(cov_df[, names(beta)]) %*% beta)
```


## 13. Convert to resistance surface

Higher predicted movement probability:
 --> lower resistance

```{r}
#| echo: FALSE
#| message: FALSE
#| warning: FALSE
cov_df$resistance <- exp(-cov_df$pred)
```

## 14. Convert back to raster

```{r}
#| echo: FALSE
#| message: FALSE
#| warning: FALSE
res_rast <- rast(
  cov_df,
  type = "xyz",
  crs = crs(covs)
)
#names(res_rast) <- "resistance"
```

## 15. Define seasonal ranges

You can:

- use MCP,
- kernel UD,
- or manually defined polygons.

Example using start/end migration locations:

```{r}
#| echo: FALSE
#| message: FALSE
#| warning: FALSE
spring_pts <- tele_sf %>%
  filter(month(timestamp) %in% 4:5)

fall_pts <- tele_sf %>%
  filter(month(timestamp) %in% 9:10)

spring_centroid <- st_centroid(st_union(spring_pts))
fall_centroid <- st_centroid(st_union(fall_pts))
```


## 16. Least-cost connectivity

Create transition object

```{r}
#| echo: FALSE
#| message: FALSE
#| warning: FALSE
tr <- transition(
  1 / raster(res_rast$resistance),
  transitionFunction = mean,
  directions = 8
)

tr <- geoCorrection(tr)
```

## 17. Compute least-cost path

```{r}
#| echo: FALSE
#| message: FALSE
#| warning: FALSE
spring_xy <- st_coordinates(spring_centroid)
fall_xy <- st_coordinates(fall_centroid)

corridor <- shortestPath(
  tr,
  origin = spring_xy,
  goal = fall_xy,
  output = "SpatialLines"
)
```


```{r}
#| echo: FALSE
#| message: FALSE
#| warning: FALSE

```


```{r}
#| echo: FALSE
#| message: FALSE
#| warning: FALSE

```

