
```{r}
#| echo: false
#| warning: false
#| message: false
library(sf)
library(amt)
library(tmap)
library(knitr)
library(tidyverse)
```

### View movement data

The table displays the first 10 rows of the caribou movement data. Attributes include caribou id, timestamp, longitude, latitude, and year for each GPS relocation.

```{r}
#| echo: TRUE
#| warning: false
#| message: false
gps <- read_csv('../../data/little_rancheria/caribou.csv') 
gps_sf <- gps |>
  st_as_sf(coords=c("longitude", "latitude"), crs=4326, remove=FALSE) |>
  mutate(year=as.factor(year))
gps
```
The map shows all 157,046 GPS relocations for the 38 individuals that were monitored between 2020-26.

```{r}
#| echo: TRUE
#| warning: false
#| message: false
tm_basemap("Esri.WorldTopoMap") +
tm_shape(gps_sf) +
  tm_dots()
```

### Number of individuals

The table below tabulates the number of individual caribou by season/migration and year.

```{r}
#| echo: TRUE
#| message: FALSE
#| warning: FALSE
row_order <- c("Early winter", "Late winter", "Calving", "Summer", "Fall rut", "Spring migration", "Fall migration")
season_tbl <- gps |>
  group_by(year, season) |>
  summarise(n_ids = n_distinct(id), .groups = "drop") |>
  pivot_wider(names_from = year, values_from = n_ids) |>
  rename(`Season/migration` = season)
migration_tbl <- gps |>
  filter(migration %in% c("Spring migration", "Fall migration")) |>
  group_by(year, migration) |>
  summarise(n_ids = n_distinct(id), .groups = "drop") |>
  pivot_wider(names_from = year, values_from = n_ids) |>
  rename(`Season/migration` = migration)
#migration_tbl <- x |>
#  filter(grepl("migration", `Season/migration`, ignore.case = TRUE)) |>
#  select(-Total)
combined_tbl <- bind_rows(season_tbl, migration_tbl) |>
  mutate(`Season/migration` = factor(`Season/migration`, levels = row_order)) |>
  arrange(`Season/migration`) |>
  mutate(Total = rowSums(pick(where(is.numeric)), na.rm = TRUE))
knitr::kable(combined_tbl)
```

### Make tracks from GPS

Tracks are the basic building blocks of movement analysis in the `amt` package. They allow us to estimate annual and seasonal ranges as well as identify migration pathways. So, we first use the `make_track` function for convert our data.frame into tracks. Note that we have to specify the crs of the data which, in this case, is lat-long or "EPSG:4326". We can then project our coordinates to the Yukon Albers projection using the `transform_coords` function.

```{r}
#| echo: TRUE
#| warning: FALSE
#| message: FALSE
trk <- make_track(gps, longitude, latitude, timestamp, id=id, year=year, season=season, crs=4326)
trk
```
