
```{r}
#| echo: false
#| warning: false
#| message: false
library(sf)
library(amt)
library(tmap)
library(knitr)
library(patchwork)
library(tidyverse)
herd_dir <- "little_rancheria"
herd_name <- "Little Rancheria"
```

## Introduction

The `r herd_name` caribou herd stradles the BC-Yukon border.

```{r}
#| echo: TRUE
#| warning: FALSE
#| message: FALSE
#| fig-width: 10
#| fig-height: 8
kdtt <- st_read("../../data/dashboard.gpkg", "kdtt", quiet=TRUE) |>
  mutate(Name="KDTT")
herd <- st_read("../../data/dashboard.gpkg", "KDTT_all_mtn_herds", quiet=TRUE) |>
  filter(HERD==herd_name)
#combined_extent <- st_union(kdtt) |>
#  st_buffer(20000) |>
#  st_bbox()
tm_basemap("Esri.WorldTopoMap") +
  tm_shape(kdtt) +
  tm_borders(col="black", lwd=2, lty="dashed") +
  tm_polygons(fill_alpha=0, col.legend = tm_legend(title = "KDTT")) +
  tm_shape(herd) +
  tm_polygons(fill="HERD", fill_alpha=0.5, col="grey",
    fill.scale = tm_scale_categorical(values="tol.rainbow"),
    fill.legend = tm_legend_hide()) +
  tm_layout(frame = FALSE) +
  tm_compass(position = tm_pos_in("left", "bottom")) +
  tm_scalebar(position = tm_pos_in("left", "bottom"))
```

### View movement data

The table displays the first 10 rows of the caribou movement data. Attributes include caribou id, timestamp, longitude, latitude, and year for each GPS relocation.

```{r}
#| echo: TRUE
#| warning: FALSE
#| message: FALSE
gps <- read_csv(paste0("../../data/",herd_dir,"/caribou.csv")) 
gps_sf <- gps |>
  st_as_sf(coords=c("longitude", "latitude"), crs=4326, remove=FALSE) |>
  mutate(year=as.factor(year))
gps
```
The map shows all 157,046 GPS relocations for the 38 individuals that were monitored between 2020-26.

```{r}
#| echo: TRUE
#| warning: FALSE
#| message: FALSE
#| fig-width: 10
#| fig-height: 8
tm_basemap("Esri.WorldTopoMap") +
tm_shape(gps_sf) +
  tm_dots()
```

### Number of individuals

The table below tabulates the number of individual caribou by season/migration and year.

```{r}
#| echo: TRUE
#| message: FALSE
#| warning: FALSE
row_order <- c("Early winter", "Late winter", "Calving", "Summer", "Fall rut", "Spring migration", "Fall migration")
season_tbl <- gps |>
  group_by(year, season) |>
  summarise(n_ids = n_distinct(id), .groups = "drop") |>
  pivot_wider(names_from = year, values_from = n_ids) |>
  rename(`Season/migration` = season)
migration_tbl <- gps |>
  filter(migration %in% c("Spring migration", "Fall migration")) |>
  group_by(year, migration) |>
  summarise(n_ids = n_distinct(id), .groups = "drop") |>
  pivot_wider(names_from = year, values_from = n_ids) |>
  rename(`Season/migration` = migration)
#migration_tbl <- x |>
#  filter(grepl("migration", `Season/migration`, ignore.case = TRUE)) |>
#  select(-Total)
combined_tbl <- bind_rows(season_tbl, migration_tbl) |>
  mutate(`Season/migration` = factor(`Season/migration`, levels = row_order)) |>
  arrange(`Season/migration`) |>
  mutate(Total = rowSums(pick(where(is.numeric)), na.rm = TRUE))
knitr::kable(combined_tbl)
```

### Make tracks from GPS

Tracks are the basic building blocks of movement analysis in the `amt` package. They allow us to estimate annual and seasonal ranges as well as identify migration pathways. So, we first use the `make_track` function for convert our data.frame into tracks. Note that we have to specify the crs of the data which, in this case, is lat-long or "EPSG:4326". We can then project our coordinates to the Yukon Albers projection using the `transform_coords` function.

```{r}
#| echo: TRUE
#| warning: FALSE
#| message: FALSE
trk <- make_track(gps, longitude, latitude, timestamp, id=id, year=year, season=season, crs=4326) |>
  transform_coords("EPSG:3578")
trk
```

### Plot sampling duration for each individual

The plot shows each individual as a horizontal segment spanning their first to last fix, ordered by total tracking duration (shortest at the bottom). A few things stand out:

- Most animals tagged in 2020-2021 have multi-year records extending to 2025-2026
- SKRATALR0057 was only tracked for a very short period around 2022
- A second cohort was collared around 2021-2022 (the 005x series)

```{r}
#| echo: TRUE
#| message: FALSE
#| warning: FALSE
gps |>
  group_by(id) |>
  summarise(start = min(timestamp), end = max(timestamp)) |>
  mutate(id = fct_rev(factor(id))) |>
  ggplot(aes(y = id)) +
  geom_segment(aes(x = start, xend = end, yend = id)) +
  geom_point(aes(x = start)) +
  geom_point(aes(x = end)) +
  labs(x = NULL, y = NULL, title = "Sampling duration by individual")
```

### Calculate movement rates

Steps are computed per individual using `steps()`, which connects consecutive GPS fixes and calculates step length (`sl_`, in metres), turning angle (`ta_`, in radians), and time elapsed (`dt_`). Speed is derived as step length divided by duration.

::: {.panel-tabset}

### Raw data

```{r}
#| echo: TRUE
#| message: FALSE
#| warning: FALSE


stps <- trk |>
  nest(data = -id) |>
  mutate(steps = map(data, steps)) |>
  select(id, steps) |>
  unnest(steps) |>
  mutate(speed_kmh = (sl_ / 1000) / as.numeric(dt_, units = "hours"))

p1 <- ggplot(stps, aes(x = sl_ / 1000)) +
  geom_histogram(bins = 60) +
  labs(x = "Step length (km)", y = "Count", title = "Step lengths")

p2 <- ggplot(stps, aes(x = speed_kmh)) +
  geom_histogram(bins = 60) +
  labs(x = "Speed (km/h)", y = "Count", title = "Speeds")

p3 <- ggplot(stps |> filter(!is.na(ta_)), aes(x = ta_)) +
  geom_histogram(bins = 36) +
  scale_x_continuous(
    breaks = c(-pi, -pi/2, 0, pi/2, pi),
    labels = c("-π", "-π/2", "0", "π/2", "π")
  ) +
  labs(x = "Turning angle (radians)", y = "Count", title = "Turning angles")

p1 / p2 / p3
```

- Step lengths and speeds are heavily right-skewed - most steps are short/slow, with a long tail of occasional fast, directed movements. This is typical for GPS data with irregular fix intervals.
- Turning angles peak near 0, indicating a tendency for directed (straight-line) movement, with secondary peaks near ±π from reversals.

### Log transformed

```{r}
#| echo: TRUE
#| message: FALSE
#| warning: FALSE


stps <- trk |>
  nest(data = -id) |>
  mutate(steps = map(data, steps)) |>
  select(id, steps) |>
  unnest(steps) |>
  mutate(speed_kmh = (sl_ / 1000) / as.numeric(dt_, units = "hours"))

p1 <- ggplot(stps, aes(x = sl_ / 1000)) +
  geom_histogram(bins = 60) +
  scale_x_log10() +
  labs(x = "Step length (km, log scale)", y = "Count", title = "Step lengths")

p2 <- ggplot(stps, aes(x = speed_kmh)) +
  geom_histogram(bins = 60) +
  scale_x_log10() +
  labs(x = "Speed (km/h, log scale)", y = "Count", title = "Speeds")

p3 <- ggplot(stps |> filter(!is.na(ta_)), aes(x = ta_)) +
  geom_histogram(bins = 36) +
  scale_x_continuous(
    breaks = c(-pi, -pi/2, 0, pi/2, pi),
    labels = c("-π", "-π/2", "0", "π/2", "π")
  ) +
  labs(x = "Turning angle (radians)", y = "Count", title = "Turning angles")

p1 / p2 / p3
```

- Both distributions look roughly log-normal on the transformed scale. 
- Step lengths are centered around ~1 km, and speeds around ~0.01–0.1 km/h, with the long tails now visible rather than collapsed against the y-axis.

:::



### Movement rates by season and migration period

Season and migration labels are joined back to steps from the original GPS data, then combined into a single `period` variable ordered chronologically through the annual cycle.

```{r}
#| echo: TRUE
#| message: FALSE
#| warning: FALSE
period_order <- c("Early winter", "Late winter", "Spring migration",
                  "Calving", "Summer", "Fall rut", "Fall migration")

stps_season <- stps |>
  left_join(
    gps |> select(id, timestamp, season, migration),
    by = c("id", "t1_" = "timestamp")
  ) |>
  mutate(
    period = if_else(!is.na(migration), migration, season),
    period = factor(period, levels = period_order)
  )

p_sl <- ggplot(stps_season, aes(x = sl_ / 1000)) +
  geom_histogram(bins = 50) +
  scale_x_log10() +
  facet_wrap(~period, nrow = 2) +
  labs(x = "Step length (km, log scale)", y = "Count", title = "Step lengths by period")

p_sp <- ggplot(stps_season, aes(x = speed_kmh)) +
  geom_histogram(bins = 50) +
  scale_x_log10() +
  facet_wrap(~period, nrow = 2) +
  labs(x = "Speed (km/h, log scale)", y = "Count", title = "Speeds by period")

p_ta <- ggplot(stps_season |> filter(!is.na(ta_)), aes(x = ta_)) +
  geom_histogram(bins = 36) +
  scale_x_continuous(
    breaks = c(-pi, 0, pi),
    labels = c("-π", "0", "π")
  ) +
  facet_wrap(~period, nrow = 2) +
  labs(x = "Turning angle (radians)", y = "Count", title = "Turning angles by period")
```

::: {.panel-tabset}

### Step lengths

```{r}
#| echo: TRUE
#| message: FALSE
#| warning: FALSE
p_sl
```

### Speeds

```{r}
#| echo: TRUE
#| message: FALSE
#| warning: FALSE
p_sp
```

### Turning angles

```{r}
#| echo: TRUE
#| message: FALSE
#| warning: FALSE
p_ta
```

:::

A few patterns worth noting from the faceted plots:

- Spring and fall migration periods show a noticeably flatter, more spread step length distribution — more long-distance steps compared to resident seasons, as you'd expect.
- Calving has very short step lengths and low speeds, consistent with restricted movement around calving sites.
- Turning angles during migration are more peaked near 0 (straighter paths), while resident seasons show more uniform/random turning.

### Median movement rates by period

The table confirms the patterns visible in the plots. Fall migration stands out as the most active period - it has the longest median steps (0.77 km) and highest speeds (0.13 km/h), while calving and late winter are the most sedentary periods - calving in particular showing both the shortest steps and lowest speeds, roughly half those of the migration periods.

```{r}
#| echo: TRUE
#| message: FALSE
#| warning: FALSE
stps_season |>
  group_by(period) |>
  summarise(
    n_steps = n(),
    median_sl_km = median(sl_ / 1000, na.rm = TRUE),
    median_speed_kmh = median(speed_kmh, na.rm = TRUE)
  ) |>
  rename(
    Period = period,
    `N steps` = n_steps,
    `Median step length (km)` = median_sl_km,
    `Median speed (km/h)` = median_speed_kmh
  ) |>
  kable(digits = 3)
```

### Movement rates by period and year

We now look at how movement rates vary by period and year. A few things are worth noting in the tables:

- The 2020 spring migration value (3.26 km) is almost certainly an artifact — there are only 25 steps recorded for that year-period combination, likely because collaring started mid-season.
- Calving step lengths have declined steadily from 0.21 km in 2021 to 0.12 km in 2025, which could be worth investigating further.
- Late winter 2026 looks low (0.12 km), but that's probably a partial year effect since data collection is ongoing.

```{r}
#| echo: TRUE
#| message: FALSE
#| warning: FALSE
rates_by_year <- stps_season |>
  mutate(year = lubridate::year(t1_)) |>
  group_by(period, year) |>
  summarise(
    median_sl_km     = median(sl_ / 1000, na.rm = TRUE),
    median_speed_kmh = median(speed_kmh, na.rm = TRUE),
    .groups = "drop"
  )
```

::: {.panel-tabset}

### Median step length (km)

```{r}
#| echo: TRUE
#| message: FALSE
#| warning: FALSE
rates_by_year |>
  select(period, year, median_sl_km) |>
  pivot_wider(names_from = year, values_from = median_sl_km) |>
  rename(Period = period) |>
  kable(digits = 3)
```

### Median speed (km/h)

```{r}
#| echo: TRUE
#| message: FALSE
#| warning: FALSE
rates_by_year |>
  select(period, year, median_speed_kmh) |>
  pivot_wider(names_from = year, values_from = median_speed_kmh) |>
  rename(Period = period) |>
  kable(digits = 3)
```

:::

### Advanced analysis

The `amt` package also provides functions for more advanced analyses, such as calculating residence time, identifying stopovers, and performing trajectory segmentation.

Here's a brief example of calculating residence time:

```R
# Calculate residence time
#residence_time <- residence_time(move_data, raster = your_raster_data)
```
