## Estimating ranges

### Overview

Estimating the boundaries of seasonal and annual caribou ranges is an important step in effective wildlife conservation and management. The why and how of estimating ranges is discussed further in the [Ranges appendix](../appendices/ranges.qmd). In this section, we first illustrate the approach for estimating individual- and population level annual and seasonal ranges. We then use the output to answer a few questions of interest. For full access to the results, we invite the user to visit the [Range Viewer dashboard](https://beaconsproject.shinyapps.io/range_viewer/).

### Walkthrough

In this section we describe the steps used to estimate individual home ranges using the KDE approach and compare it to another commonly used method, the minimum convex hull (MCP). We illustrate the approach for one individual caribou.

```{r}
#| echo: false
#| warning: false
#| message: false
library(sf)
library(amt)
library(tmap)
library(knitr)
library(patchwork)
library(tidyverse)
```

#### Step 1. Read locations

```{r}
#| echo: false
#| warning: false
#| message: false
#| fig-width: 10
#| fig-height: 8
gps <- read_csv('../../data/little_rancheria/caribou.csv')
gps1 <- gps |> filter(id=="SKRATALR0043" & year==2022)
gps1_sf <- gps1 |>
  st_as_sf(coords=c("longitude", "latitude"), crs=4326, remove=FALSE) |>
  mutate(year=as.factor(year))
tm_basemap("Esri.WorldTopoMap") +
  tm_shape(gps1_sf) +
  tm_dots(fill="red")
```

#### Step 2. Make tracks

Tracks are the basic building blocks of movement analysis in the `amt` package. So, we first use the `make_track` function for convert our data.frame into tracks. Note that we have to specify the crs of the data which, in this case, is lat-long or "EPSG:4326". We can then project our coordinates to the Yukon Albers projection using the `transform_coords` function.

```{r}
#| echo: false
#| warning: false
#| message: false
#| fig-width: 10
#| fig-height: 8
trk <- make_track(gps, longitude, latitude, timestamp, id=id, year=year, season=season, crs=4326)
trk1 <- make_track(gps1, longitude, latitude, timestamp, id=id, year=year, season=season, crs=4326)
trk1_sf <- trk1 |>
  group_by(id, year) |>
  st_as_sf(coords=c("x_", "y_"), crs=4326) |>
  summarize(do_union=FALSE) |>
  st_cast("LINESTRING") |>
  st_make_valid()
tm_basemap("Esri.WorldTopoMap") +
  tm_shape(trk1_sf) +
  tm_lines(lwd=2) +
  tm_shape(gps1_sf) +
  tm_dots(fill="red")
```

#### Step 3. Estimate ranges

**Geometric estimators: MCP**

Geometric estimators such as minimum convex polygons(MCP) only rely on the observed locations and can only estimate the occurrence distribution (i.e., where the animal was during the study period). MCPs can be constructed with multiple levels using the function `hr_mcp()` from the `amt` package.

Estimated home range area for each level:

```{r}
#| echo: false
#| warning: false
#| message: false
mcps <- hr_mcp(trk1, levels = c(0.5, 0.75, 0.95, 1))
hr_area(mcps, units = TRUE) %>% 
  mutate(area = units::set_units(area, "km^2"))
```

MCPs can be viewed as isopleths.

```{r}
#| echo: false
#| warning: false
#| message: false
#| fig-width: 10
#| fig-height: 8
iso1 <- hr_isopleths(mcps) |>
  mutate(level = factor(level))
tm_basemap("Esri.WorldTopoMap") +
  tm_shape(gps1_sf) +
  tm_dots() +
  tm_shape(iso1) +
  tm_borders(col="level", col.scale=tm_scale(values=c("red","darkgreen","blue","black")), lwd=2, col.legend = tm_legend_hide()) +
  tm_add_legend(title="MCP Level", labels=c("50%","75%","95%","100%"), col=c("red","darkgreen","blue","black"), 
    fill_alpha=0, position = tm_pos_in(pos.v = "bottom", pos.h="right")) 
```

**Probabilistic estimators (KDE)**

Probabilistic estimators (e.g., KDEs) are based on an underlying model for the utilization distribution (UD). They can estimate the occurrence distribution or the range distribution (where the animal could be found at any future time if its movement process remains the same) depending on the underlying model. Can be represented as a hull by outlining a cumulative probability of use. Fitting a KDE to telemetry data is the same problem as with MCPs, but in 2 dimensions. Choosing a bandwidth is important to the final result, and there are multiple possible methods for doing so. Here, we will focus on the reference bandwidth, whic is the default.

```{r}
#| echo: false
#| warning: false
#| message: false
kde1 <- hr_kde(trk1, levels = c(0.5, 0.95))
hr_area(kde1)
```

Unlike MCPs, KDEs include two additional raster objects:

* 'ud' -- the rasterized utilization distribution
* 'trast' -- the template raster used to rasterize the UD

Here we make a plot with the 50% and 95% isopleths drawn over the UD raster.

```{r}
#| echo: false
#| warning: false
#| message: false
#| fig-width: 10
#| fig-height: 8
kde1 <- hr_kde(trk1, levels = c(0.5, 0.95))
kde1i <- kde1 |>
  hr_isopleths()
tm_basemap("Esri.WorldTopoMap") +
  tm_shape(kde1i) +
  tm_borders(col="black", lwd=2, lty="dashed") +
  tm_polygons( 
    fill="level", fill.scale=tm_scale(values=c("yellow","darkgreen")),
    fill_alpha=0.5, fill.legend = tm_legend_hide()) +
  tm_add_legend(
    title = "KDE", labels = c("Core (50%)", "Overall (95%)"), fill=c("yellow","darkgreen"), fill_alpha=0.75, position = tm_pos_in(pos.v = "bottom", pos.h="right"))
```

#### Step 4. Compare estimates

```{r}
#| echo: false
#| warning: false
#| message: false
#| fig-width: 10
#| fig-height: 8
mcp1 <- hr_mcp(trk1, levels = c(0.5, 0.95))
mcp1i <- mcp1 |> hr_isopleths()
kde1 <- hr_kde(trk1, levels = c(0.5, 0.95))
kde1i <- kde1 |> hr_isopleths()
tm_basemap("Esri.WorldTopoMap") +
  tm_shape(kde1i) +
  tm_borders(col="black", lwd=2, lty="dashed") +
  tm_polygons( 
    fill="level", fill.scale=tm_scale(values=c("yellow","darkgreen")),
    fill_alpha=0.5, fill.legend = tm_legend_hide()) +
  tm_add_legend(
    title = "KDE", labels = c("Core (50%)", "Overall (95%)"), fill=c("yellow","darkgreen"), fill_alpha=0.75, position = tm_pos_in(pos.v = "bottom", pos.h="right")) +
  tm_shape(mcp1i) +
  tm_borders(col="level", col.scale=tm_scale(values=c("red","blue")), lwd=2, lty="solid", col.legend = tm_legend_hide()) +
  tm_add_legend(title="MCP", labels=c("Core (50%)","Overall (95%)"), col=c("red","blue"), fill_alpha=0, position = tm_pos_in(pos.v = "bottom", pos.h="right")) +
  tm_shape(gps1_sf) +
  tm_dots(col="black", size=0.1) +
  tm_compass(position = tm_pos_in("right", "top")) +
  tm_scalebar(position = tm_pos_in("left", "top"))
```
### Measure overlap between approaches

We can now compare the overlap in estimated ranges between the two methods:

```{r}
#| warning: false
#| message: false
#hr_isopleths(mcp1)
ovr1 <- hr_overlap(mcp1, kde1)
ovr2 <- hr_overlap(kde1, mcp1)
tibble(
  Overlap = c("MCP estimate overlaps KDE estimate by:", "KDE estimate overlaps MCP estimate by:"),
  `Overall area (95%)` = c(round(ovr1$overlap[1]*100,1), round(ovr2$overlap[1]*100,1)),
  `Overall area (50%)` = c(round(ovr1$overlap[2]*100,1), round(ovr2$overlap[2]*100,1))) |>
  kable()
```
