---
title: "Seasonal RSFs"
---

```{r}
#| echo: false
#| warning: false
#| message: false
library(terra)
library(sf)
library(tmap)
library(tidyterra)
library(glmmTMB)
library(knitr)
herd_dir <- "little_rancheria"
herd_name <- "Little Rancheria"
source("../R/rsf_map.R")
output_dir <- "../../models/little_rancheria/"

# Bigger, full-width map figures.
knitr::opts_chunk$set(fig.width = 10, fig.height = 8, out.width = "100%")

# Seasons shown in Results below. "Annual" is also a valid season1 value in
# little_rancheria_rsf_workflow.R (pooling all seasons) but is intentionally
# left out of the season-by-season Results.
seasons <- c("Calving", "Summer", "Fall rut",
             "Early winter", "Late winter")

term_labels <- c(
  elevation_z = "elevation",
  tri_z       = "terrain ruggedness (TRI)",
  tpi_z       = "terrain position (TPI)",
  northness_z = "northness",
  eastness_z  = "eastness"
)

# Pull a tidy fixed-effects table out of a fitted glmmTMB model without
# depending on broom.mixed (not part of the workflow's package list).
tidy_fixed <- function(model) {
  cc <- summary(model)$coefficients$cond
  df <- data.frame(Term = rownames(cc), round(cc, 3), row.names = NULL)
  names(df) <- c("Term", "Estimate", "Std. Error", "z value", "p value")
  df
}

# Random-slope SDs by id (elevation_z, tri_z). The (1 | id) random-intercept
# variance is fixed at log(1e3) by design (Muff et al. 2020 weighting
# approach, see Methods) rather than estimated, so it is not a meaningful
# variance component and is dropped by the caller before display.
tidy_ranef <- function(model) {
  vc <- VarCorr(model)$cond
  rows <- Map(function(v, grp) {
    sd <- attr(v, "stddev")
    data.frame(Group = grp, Term = names(sd), SD = round(unname(sd), 3),
               row.names = NULL)
  }, vc, names(vc))
  do.call(rbind, rows)
}

# Short, data-driven interpretation of a model's fixed effects, ordered by
# strength of evidence (|z value|). Only fixed effects are discussed here,
# since the accompanying map is a fixed-effects-only (re.form = NA)
# prediction (see Methods).
interpret_fixed <- function(model) {
  fx <- tidy_fixed(model)
  fx <- fx[fx$Term != "(Intercept)", ]
  fx$Label     <- term_labels[fx$Term]
  fx$Direction <- ifelse(fx$Estimate > 0, "selection for", "avoidance of")
  fx$Sig       <- fx$`p value` < 0.05
  fx <- fx[order(-abs(fx$`z value`)), ]

  fmt_p <- function(p) if (p < 0.001) "p < 0.001" else paste0("p = ", round(p, 3))

  sig    <- fx[fx$Sig, ]
  nonsig <- fx[!fx$Sig, ]

  if (nrow(sig) > 0) {
    parts <- sprintf("%s (%s, β = %.2f, %s)",
                      sig$Label, sig$Direction, sig$Estimate,
                      sapply(sig$`p value`, fmt_p))
    sig_txt <- paste0("Strongest evidence of selection was for ",
                       paste(parts, collapse = "; "), ".")
  } else {
    sig_txt <- paste("None of the topographic covariates showed",
                      "statistically clear selection (p < 0.05) at this",
                      "order/season -- the predicted surface reflects a",
                      "comparatively flat response.")
  }

  nonsig_txt <- if (nrow(nonsig) > 0 && nrow(sig) > 0) {
    paste0(" ", paste(nonsig$Label, collapse = ", "),
           " showed no clear evidence of selection.")
  } else {
    ""
  }

  cat(sig_txt, nonsig_txt, "\n")
}

# Load a model/raster if it exists, otherwise note that the season/order
# hasn't been (re-)run yet, so a partial set of outputs doesn't break the
# whole render.
load_model <- function(path) {
  if (!file.exists(path)) {
    cat("Not yet available:", path, "\n")
    return(invisible(NULL))
  }
  readRDS(path)
}

load_raster <- function(path) {
  if (!file.exists(path)) {
    cat("Not yet available:", path, "\n")
    return(invisible(NULL))
  }
  terra::rast(path)
}

load_vector <- function(path) {
  if (!file.exists(path)) {
    cat("Not yet available:", path, "\n")
    return(invisible(NULL))
  }
  sf::st_read(path, quiet = TRUE)
}
```

# Introduction

The Little Rancheria caribou herd occupies a range spanning several
seasonal phases (calving, summer, fall rut, early winter, late winter),
each associated with potentially distinct habitat requirements. Resource
selection functions (RSFs) provide a statistical framework for
quantifying how strongly animals select landscape features relative to
their availability, and are commonly estimated at multiple spatial
scales, or "orders of selection" (Johnson 1980): **2nd-order** selection
describes placement of an individual's home range within the broader
study area, while **3rd-order** selection describes use of locations
within that home range.

This report summarizes seasonal 2nd- and 3rd-order RSFs fit for the
Little Rancheria herd from GPS telemetry data and terrain covariates
derived from a regional digital elevation model (DEM).

## Objectives

- Characterize seasonal patterns of topographic habitat selection (elevation,
  ruggedness, terrain position, and slope orientation) at both the
  home-range (2nd-order) and within-home-range (3rd-order) scales.
- Fit weighted mixed-effects RSFs that account for repeated sampling of
  individuals and the used:available design.
- Generate spatial predictions of relative probability of use across the
  herd range for each season, to support habitat mapping and management
  applications.

# Methods

The full analysis workflow is implemented in
`little_rancheria_rsf_workflow.R`, run once per season by setting the
`season1` variable (`"Calving"`, `"Summer"`, `"Fall rut"`,
`"Early winter"`, `"Late winter"`, or `"Annual"`) and re-running the
script; outputs are written to a folder named for that season. The
sections below summarize the workflow; code shown is illustrative
(`eval: false`) rather than re-executed in this report.

## Movement data and study area

GPS telemetry fixes (`caribou.csv`) were filtered to the season of
interest using a `phase` column, parsed to timestamps, and converted to
`amt` tracks per individual (`make_track()`), reprojected to match the
CRS of a regional DEM (`mrdem.tif`).

```{r}
#| label: methods-tracks
#| eval: false

trk_all <- caribou_raw |>
  make_track(x, y, t, id = id, crs = 4326) |>
  transform_coords(crs_to = target_crs)
```

## Terrain covariates

The DEM was cropped to the telemetry extent plus a 5 km buffer (to keep
raster processing tractable and ensure coverage for available points near
the range edge), then used to derive five topographic covariates with
`terra::terrain()`:

- **elevation** (m)
- **TRI** (terrain ruggedness index)
- **TPI** (topographic position index)
- **northness** and **eastness** (cosine/sine of aspect, used in place of
  circular aspect degrees)

Slope and roughness were dropped as redundant with TRI (all three capture
local ruggedness).

## Home ranges and availability sampling

Home ranges were estimated per individual with both MCP and KDE
(`amt::hr_mcp()`, `amt::hr_kde()`; 95% level) as a sanity check on area
estimates. Autocorrelated KDE (aKDE) was deliberately not used, since it
assumes range residency that does not hold for this herd.

Used-available data were generated at two scales, with a 1:10 used:available
ratio:

- **3rd order**: available points drawn within each individual's own KDE
  home range (`amt::random_points()`).
- **2nd order**: available points drawn from the union of all individuals'
  MCPs (a stand-in herd/study-area extent).

Covariates were extracted at all used and available points and
standardized (z-scores) prior to model fitting. Collinearity among the
five (unstandardized) covariates was checked separately for each order's
dataset with `usdm::vifstep()` (threshold VIF = 5).

## RSF models

Models followed the weighted mixed-effects logistic regression approach of
Muff, Signer & Fieberg (2020): available (pseudo-absence) points were
down-weighted by assigning them a weight of 1000 relative to used points
(weight = 1), and the variance of the individual-level random intercept
was fixed at a large value (`log(1e3)`) so that it absorbs the
used:available sampling ratio rather than competing with the fixed-effect
slope estimates. Random slopes for elevation and TRI were left free to
vary by individual.

```{r}
#| label: methods-formula
#| eval: false

rsf_formula <- paste(
  "elevation_z + tri_z + tpi_z + northness_z + eastness_z",
  "+ (1 | id)",
  "+ (0 + elevation_z | id)",
  "+ (0 + tri_z | id)"
)

m_2nd <- fit_weighted_rsf(second_order_dat, rsf_formula)
m_3rd <- fit_weighted_rsf(third_order_dat, rsf_formula)
```

A full-model formula (all five topographic covariates) was fit for both
orders in every season; the script also includes a `model_selection()`
helper comparing this full model against reduced topo-only candidate
models via AICc, for use once additional covariate layers (e.g.
disturbance features) are added.

## Spatial prediction

The fitted fixed effects from *each* order's model were used to predict a
population-level (fixed-effects-only, `re.form = NA`) RSF surface across
the cropped landscape. Each prediction re-derives the z-standardization
used to fit that model -- 2nd- and 3rd-order used-available data were
standardized separately (Section 6 above), so the 2nd- and 3rd-order
covariate means/sds differ and each prediction stack is built from its own
model's training data. Rather than calling `terra::predict()` directly on
the `glmmTMB` object (which can build an oversized in-memory model matrix
per raster block and trigger `std::bad_alloc` on large rasters), the
linear predictor was computed by hand as raster algebra from the
fixed-effect coefficients and covariate rasters, then back-transformed to
the probability scale with the inverse logit. The results for each season
were saved as `little_rancheria_rsf_2ndorder.tif` and
`little_rancheria_rsf_3rdorder.tif`.

# Results

Results are split into two sections below, one per order of selection.
Within each section, seasons are shown as tabs; each tab gives the fixed-
effects summary of that order's model (`m_2nd.RDS` or `m_3rd.RDS`),
the random-slope SDs (among-individual variation in selection strength),
the corresponding predictive RSF surface (`little_rancheria_rsf_2ndorder.tif`
or `little_rancheria_rsf_3rdorder.tif`) with that season's GPS relocations
and 100% MCP overlaid, and a short interpretation of that map.
Seasons/orders that haven't been (re-)run yet are flagged rather than
breaking the render. The "Annual" model is fit separately by the workflow
script but omitted from these season-by-season results.

## 2nd-order results

2nd-order models describe selection of home-range placement within the
broader herd/study-area extent.

::: {.panel-tabset}

### Calving

```{r}
#| label: calving-2nd-model

m_2nd <- load_model(paste0(output_dir,"Calving/m_2nd.RDS"))
if (!is.null(m_2nd)) kable(tidy_fixed(m_2nd), caption = "Calving – 2nd-order fixed effects")
```

```{r}
#| label: calving-2nd-ranef

if (!is.null(m_2nd)) {
  rt <- tidy_ranef(m_2nd)
  rt <- rt[rt$Term != "(Intercept)", ]
  kable(rt, caption = "Calving – 2nd-order random-slope SDs (by id); random-intercept variance is fixed by design and omitted")
}
```

```{r}
#| label: calving-2nd-map

r_2nd <- load_raster(paste0(output_dir,"Calving/little_rancheria_rsf_2ndorder.tif"))
gps_pts <- load_vector(paste0(output_dir,"Calving/gps_points.gpkg"))
mcp_100 <- load_vector(paste0(output_dir,"Calving/mcp_100.gpkg"))

if (!is.null(r_2nd)) {
  m <- rsf_map(r_2nd, title = "Calving – 2nd-order RSF")
  if (!is.null(mcp_100)) m <- m + tm_shape(mcp_100) + tm_borders(col = "black", lwd = 2)
  if (!is.null(gps_pts)) m <- m + tm_shape(gps_pts) + tm_dots(col = "black", size = 0.02, alpha = 0.5)
  m
}
```

```{r}
#| label: calving-2nd-interp
#| output: asis

if (!is.null(m_2nd)) interpret_fixed(m_2nd)
```

### Summer

```{r}
#| label: summer-2nd-model

m_2nd <- load_model(paste0(output_dir,"Summer/m_2nd.RDS"))
if (!is.null(m_2nd)) kable(tidy_fixed(m_2nd), caption = "Summer – 2nd-order fixed effects")
```

```{r}
#| label: summer-2nd-ranef

if (!is.null(m_2nd)) {
  rt <- tidy_ranef(m_2nd)
  rt <- rt[rt$Term != "(Intercept)", ]
  kable(rt, caption = "Summer – 2nd-order random-slope SDs (by id); random-intercept variance is fixed by design and omitted")
}
```

```{r}
#| label: summer-2nd-map

r_2nd <- load_raster(paste0(output_dir,"Summer/little_rancheria_rsf_2ndorder.tif"))
gps_pts <- load_vector(paste0(output_dir,"Summer/gps_points.gpkg"))
mcp_100 <- load_vector(paste0(output_dir,"Summer/mcp_100.gpkg"))

if (!is.null(r_2nd)) {
  m <- rsf_map(r_2nd, title = "Summer – 2nd-order RSF")
  if (!is.null(mcp_100)) m <- m + tm_shape(mcp_100) + tm_borders(col = "black", lwd = 2)
  if (!is.null(gps_pts)) m <- m + tm_shape(gps_pts) + tm_dots(col = "black", size = 0.02, alpha = 0.5)
  m
}
```

```{r}
#| label: summer-2nd-interp
#| output: asis

if (!is.null(m_2nd)) interpret_fixed(m_2nd)
```

### Fall rut

```{r}
#| label: fallrut-2nd-model

m_2nd <- load_model(paste0(output_dir,"Fall rut/m_2nd.RDS"))
if (!is.null(m_2nd)) kable(tidy_fixed(m_2nd), caption = "Fall rut – 2nd-order fixed effects")
```

```{r}
#| label: fallrut-2nd-ranef

if (!is.null(m_2nd)) {
  rt <- tidy_ranef(m_2nd)
  rt <- rt[rt$Term != "(Intercept)", ]
  kable(rt, caption = "Fall rut – 2nd-order random-slope SDs (by id); random-intercept variance is fixed by design and omitted")
}
```

```{r}
#| label: fallrut-2nd-map

r_2nd <- load_raster(paste0(output_dir,"Fall rut/little_rancheria_rsf_2ndorder.tif"))
gps_pts <- load_vector(paste0(output_dir,"Fall rut/gps_points.gpkg"))
mcp_100 <- load_vector(paste0(output_dir,"Fall rut/mcp_100.gpkg"))

if (!is.null(r_2nd)) {
  m <- rsf_map(r_2nd, title = "Fall rut – 2nd-order RSF")
  if (!is.null(mcp_100)) m <- m + tm_shape(mcp_100) + tm_borders(col = "black", lwd = 2)
  if (!is.null(gps_pts)) m <- m + tm_shape(gps_pts) + tm_dots(col = "black", size = 0.02, alpha = 0.5)
  m
}
```

```{r}
#| label: fallrut-2nd-interp
#| output: asis

if (!is.null(m_2nd)) interpret_fixed(m_2nd)
```

### Early winter

```{r}
#| label: earlywinter-2nd-model

m_2nd <- load_model(paste0(output_dir,"Early winter/m_2nd.RDS"))
if (!is.null(m_2nd)) kable(tidy_fixed(m_2nd), caption = "Early winter – 2nd-order fixed effects")
```

```{r}
#| label: earlywinter-2nd-ranef

if (!is.null(m_2nd)) {
  rt <- tidy_ranef(m_2nd)
  rt <- rt[rt$Term != "(Intercept)", ]
  kable(rt, caption = "Early winter – 2nd-order random-slope SDs (by id); random-intercept variance is fixed by design and omitted")
}
```

```{r}
#| label: earlywinter-2nd-map

r_2nd <- load_raster(paste0(output_dir,"Early winter/little_rancheria_rsf_2ndorder.tif"))
gps_pts <- load_vector(paste0(output_dir,"Early winter/gps_points.gpkg"))
mcp_100 <- load_vector(paste0(output_dir,"Early winter/mcp_100.gpkg"))

if (!is.null(r_2nd)) {
  m <- rsf_map(r_2nd, title = "Early winter – 2nd-order RSF")
  if (!is.null(mcp_100)) m <- m + tm_shape(mcp_100) + tm_borders(col = "black", lwd = 2)
  if (!is.null(gps_pts)) m <- m + tm_shape(gps_pts) + tm_dots(col = "black", size = 0.02, alpha = 0.5)
  m
}
```

```{r}
#| label: earlywinter-2nd-interp
#| output: asis

if (!is.null(m_2nd)) interpret_fixed(m_2nd)
```

### Late winter

```{r}
#| label: latewinter-2nd-model

m_2nd <- load_model(paste0(output_dir,"Late winter/m_2nd.RDS"))
if (!is.null(m_2nd)) kable(tidy_fixed(m_2nd), caption = "Late winter – 2nd-order fixed effects")
```

```{r}
#| label: latewinter-2nd-ranef

if (!is.null(m_2nd)) {
  rt <- tidy_ranef(m_2nd)
  rt <- rt[rt$Term != "(Intercept)", ]
  kable(rt, caption = "Late winter – 2nd-order random-slope SDs (by id); random-intercept variance is fixed by design and omitted")
}
```

```{r}
#| label: latewinter-2nd-map

r_2nd <- load_raster(paste0(output_dir,"Late winter/little_rancheria_rsf_2ndorder.tif"))
gps_pts <- load_vector(paste0(output_dir,"Late winter/gps_points.gpkg"))
mcp_100 <- load_vector(paste0(output_dir,"Late winter/mcp_100.gpkg"))

if (!is.null(r_2nd)) {
  m <- rsf_map(r_2nd, title = "Late winter – 2nd-order RSF")
  if (!is.null(mcp_100)) m <- m + tm_shape(mcp_100) + tm_borders(col = "black", lwd = 2)
  if (!is.null(gps_pts)) m <- m + tm_shape(gps_pts) + tm_dots(col = "black", size = 0.02, alpha = 0.5)
  m
}
```

```{r}
#| label: latewinter-2nd-interp
#| output: asis

if (!is.null(m_2nd)) interpret_fixed(m_2nd)
```

:::

## 3rd-order results

3rd-order models describe selection of locations within each individual's
own home range.

::: {.panel-tabset}

### Calving

```{r}
#| label: calving-3rd-model

m_3rd <- load_model(paste0(output_dir,"Calving/m_3rd.RDS"))
if (!is.null(m_3rd)) kable(tidy_fixed(m_3rd), caption = "Calving – 3rd-order fixed effects")
```

```{r}
#| label: calving-3rd-ranef

if (!is.null(m_3rd)) {
  rt <- tidy_ranef(m_3rd)
  rt <- rt[rt$Term != "(Intercept)", ]
  kable(rt, caption = "Calving – 3rd-order random-slope SDs (by id); random-intercept variance is fixed by design and omitted")
}
```

```{r}
#| label: calving-3rd-map

r_3rd <- load_raster(paste0(output_dir,"Calving/little_rancheria_rsf_3rdorder.tif"))
gps_pts <- load_vector(paste0(output_dir,"Calving/gps_points.gpkg"))
mcp_100 <- load_vector(paste0(output_dir,"Calving/mcp_100.gpkg"))

if (!is.null(r_3rd)) {
  m <- rsf_map(r_3rd, title = "Calving – 3rd-order RSF")
  if (!is.null(mcp_100)) m <- m + tm_shape(mcp_100) + tm_borders(col = "black", lwd = 2)
  if (!is.null(gps_pts)) m <- m + tm_shape(gps_pts) + tm_dots(col = "black", size = 0.02, alpha = 0.5)
  m
}
```

```{r}
#| label: calving-3rd-interp
#| output: asis

if (!is.null(m_3rd)) interpret_fixed(m_3rd)
```

### Summer

```{r}
#| label: summer-3rd-model

m_3rd <- load_model(paste0(output_dir,"Summer/m_3rd.RDS"))
if (!is.null(m_3rd)) kable(tidy_fixed(m_3rd), caption = "Summer – 3rd-order fixed effects")
```

```{r}
#| label: summer-3rd-ranef

if (!is.null(m_3rd)) {
  rt <- tidy_ranef(m_3rd)
  rt <- rt[rt$Term != "(Intercept)", ]
  kable(rt, caption = "Summer – 3rd-order random-slope SDs (by id); random-intercept variance is fixed by design and omitted")
}
```

```{r}
#| label: summer-3rd-map

r_3rd <- load_raster(paste0(output_dir,"Summer/little_rancheria_rsf_3rdorder.tif"))
gps_pts <- load_vector(paste0(output_dir,"Summer/gps_points.gpkg"))
mcp_100 <- load_vector(paste0(output_dir,"Summer/mcp_100.gpkg"))

if (!is.null(r_3rd)) {
  m <- rsf_map(r_3rd, title = "Summer – 3rd-order RSF")
  if (!is.null(mcp_100)) m <- m + tm_shape(mcp_100) + tm_borders(col = "black", lwd = 2)
  if (!is.null(gps_pts)) m <- m + tm_shape(gps_pts) + tm_dots(col = "black", size = 0.02, alpha = 0.5)
  m
}
```

```{r}
#| label: summer-3rd-interp
#| output: asis

if (!is.null(m_3rd)) interpret_fixed(m_3rd)
```

### Fall rut

```{r}
#| label: fallrut-3rd-model

m_3rd <- load_model(paste0(output_dir,"Fall rut/m_3rd.RDS"))
if (!is.null(m_3rd)) kable(tidy_fixed(m_3rd), caption = "Fall rut – 3rd-order fixed effects")
```

```{r}
#| label: fallrut-3rd-ranef

if (!is.null(m_3rd)) {
  rt <- tidy_ranef(m_3rd)
  rt <- rt[rt$Term != "(Intercept)", ]
  kable(rt, caption = "Fall rut – 3rd-order random-slope SDs (by id); random-intercept variance is fixed by design and omitted")
}
```

```{r}
#| label: fallrut-3rd-map

r_3rd <- load_raster(paste0(output_dir,"Fall rut/little_rancheria_rsf_3rdorder.tif"))
gps_pts <- load_vector(paste0(output_dir,"Fall rut/gps_points.gpkg"))
mcp_100 <- load_vector(paste0(output_dir,"Fall rut/mcp_100.gpkg"))

if (!is.null(r_3rd)) {
  m <- rsf_map(r_3rd, title = "Fall rut – 3rd-order RSF")
  if (!is.null(mcp_100)) m <- m + tm_shape(mcp_100) + tm_borders(col = "black", lwd = 2)
  if (!is.null(gps_pts)) m <- m + tm_shape(gps_pts) + tm_dots(col = "black", size = 0.02, alpha = 0.5)
  m
}
```

```{r}
#| label: fallrut-3rd-interp
#| output: asis

if (!is.null(m_3rd)) interpret_fixed(m_3rd)
```

### Early winter

```{r}
#| label: earlywinter-3rd-model

m_3rd <- load_model(paste0(output_dir,"Early winter/m_3rd.RDS"))
if (!is.null(m_3rd)) kable(tidy_fixed(m_3rd), caption = "Early winter – 3rd-order fixed effects")
```

```{r}
#| label: earlywinter-3rd-ranef

if (!is.null(m_3rd)) {
  rt <- tidy_ranef(m_3rd)
  rt <- rt[rt$Term != "(Intercept)", ]
  kable(rt, caption = "Early winter – 3rd-order random-slope SDs (by id); random-intercept variance is fixed by design and omitted")
}
```

```{r}
#| label: earlywinter-3rd-map

r_3rd <- load_raster(paste0(output_dir,"Early winter/little_rancheria_rsf_3rdorder.tif"))
gps_pts <- load_vector(paste0(output_dir,"Early winter/gps_points.gpkg"))
mcp_100 <- load_vector(paste0(output_dir,"Early winter/mcp_100.gpkg"))

if (!is.null(r_3rd)) {
  m <- rsf_map(r_3rd, title = "Early winter – 3rd-order RSF")
  if (!is.null(mcp_100)) m <- m + tm_shape(mcp_100) + tm_borders(col = "black", lwd = 2)
  if (!is.null(gps_pts)) m <- m + tm_shape(gps_pts) + tm_dots(col = "black", size = 0.02, alpha = 0.5)
  m
}
```

```{r}
#| label: earlywinter-3rd-interp
#| output: asis

if (!is.null(m_3rd)) interpret_fixed(m_3rd)
```

### Late winter

```{r}
#| label: latewinter-3rd-model

m_3rd <- load_model(paste0(output_dir,"Late winter/m_3rd.RDS"))
if (!is.null(m_3rd)) kable(tidy_fixed(m_3rd), caption = "Late winter – 3rd-order fixed effects")
```

```{r}
#| label: latewinter-3rd-ranef

if (!is.null(m_3rd)) {
  rt <- tidy_ranef(m_3rd)
  rt <- rt[rt$Term != "(Intercept)", ]
  kable(rt, caption = "Late winter – 3rd-order random-slope SDs (by id); random-intercept variance is fixed by design and omitted")
}
```

```{r}
#| label: latewinter-3rd-map

r_3rd <- load_raster(paste0(output_dir,"Late winter/little_rancheria_rsf_3rdorder.tif"))
gps_pts <- load_vector(paste0(output_dir,"Late winter/gps_points.gpkg"))
mcp_100 <- load_vector(paste0(output_dir,"Late winter/mcp_100.gpkg"))

if (!is.null(r_3rd)) {
  m <- rsf_map(r_3rd, title = "Late winter – 3rd-order RSF")
  if (!is.null(mcp_100)) m <- m + tm_shape(mcp_100) + tm_borders(col = "black", lwd = 2)
  if (!is.null(gps_pts)) m <- m + tm_shape(gps_pts) + tm_dots(col = "black", size = 0.02, alpha = 0.5)
  m
}
```

```{r}
#| label: latewinter-3rd-interp
#| output: asis

if (!is.null(m_3rd)) interpret_fixed(m_3rd)
```

:::

# Discussion and conclusions

Seasonal RSFs provide a way to characterize how topographic selection by
the Little Rancheria herd shifts across the annual cycle, at both the
home-range placement (2nd-order) and within-home-range (3rd-order)
scales. Comparing coefficient signs and magnitudes for elevation, terrain
ruggedness (TRI), terrain position (TPI), and slope orientation
(northness/eastness) across seasons can highlight, for example, whether
the herd selects for higher/more rugged terrain during calving (often
associated with predator avoidance) versus lower, more sheltered terrain
in late winter.

A few caveats are worth keeping in mind when interpreting these results:

- **Topography-only covariate set.** The current models include only
  terrain-derived covariates. Anthropogenic and vegetation covariates
  (e.g., roads, cutblocks, seismic lines, land cover) are not yet
  included, and adding them may change both the relative importance of
  terrain variables and the collinearity structure (Section 6b of the
  workflow script should be re-run against any expanded covariate set).
- **Used:available ratio.** A 1:10 ratio with a corresponding weight of
  1000 on available points was used throughout; sensitivity of the
  fitted coefficients to this ratio has not yet been formally tested.
- **Model selection.** The candidate model comparisons in the workflow
  script are illustrative reduced sets; a more complete hypothesis-driven
  candidate set (and validation, e.g., k-fold cross-validation of
  predicted RSF surfaces) would strengthen confidence in the selected
  model structure for each season.

Overall, the workflow produces a consistent, reproducible basis for
seasonal habitat mapping across the herd's range, and is structured so
that additional covariate layers or candidate model sets can be
incorporated season-by-season without changing the core modeling
approach.
