---
title: Seasonal Migration Pathways
date: 2026-04-29
author: Pierre Vernier
---
  
```{r}
#| echo: TRUE
#| warning: FALSE
#| message: FALSE
library(sf)
library(tidyverse)
library(patchwork)
library(leaflet)
library(htmlwidgets)
library(htmltools)
library(RColorBrewer)
herd <- "little_rancheria"
```

```{r}
#| echo: FALSE
#| warning: FALSE
#| message: FALSE
# Shared year colour palette (used by both seasons)
make_season_map <- function(data, select_id, label) {
  years_all  <- sort(unique(data$year))
  n_col      <- max(3, length(years_all))
  yr_colors  <- brewer.pal(n_col, "Dark2")[seq_along(years_all)]
  year_pal   <- colorFactor(yr_colors, domain = as.character(years_all))

  ids <- sort(unique(data$id))

  m <- leaflet(options = leafletOptions(minZoom = 4)) |>
    addProviderTiles(providers$CartoDB.Positron)

  for (id_val in ids) {
    id_data <- data |> filter(id == id_val) |> arrange(year, timestamp)
    for (yr in sort(unique(id_data$year))) {
      yr_data <- id_data |> filter(year == yr)
      m <- m |> addPolylines(
        lng     = yr_data$longitude,
        lat     = yr_data$latitude,
        color   = year_pal(as.character(yr)),
        weight  = 2, opacity = 0.85,
        group   = as.character(id_val),
        label   = paste("Year:", yr)
      )
    }
  }

  # Show only the first ID on load
  for (id_val in ids[-1]) {
    m <- m |> hideGroup(as.character(id_val))
  }

  m <- m |> addLegend(
    position = "bottomright",
    colors   = yr_colors,
    labels   = as.character(years_all),
    title    = "Year",
    opacity  = 0.85
  )

  ids_json <- jsonlite::toJSON(as.character(ids))
  m <- m |> onRender(sprintf("
    function(el, x) {
      var map  = this;
      var ids  = %s;
      var sel  = document.getElementById('%s');
      sel.addEventListener('change', function() {
        var chosen = this.value;
        ids.forEach(function(id) {
          if (id === chosen) {
            map.layerManager.showGroup(id);
          } else {
            map.layerManager.hideGroup(id);
          }
        });
      });
    }
  ", ids_json, select_id))

  browsable(tagList(
    tags$div(
      style = "margin: 6px 0 4px 0;",
      tags$label(
        label,
        style = "font-weight:600; margin-right:8px;"
      ),
      tags$select(
        id    = select_id,
        style = "font-size:13px; padding:2px 6px;",
        lapply(ids, function(id_val) tags$option(value = id_val, id_val))
      )
    ),
    m
  ))
}
```

### Read caribou movement (gps) data

```{r}
#| echo: TRUE
#| warning: FALSE
#| message: FALSE
y <- st_read(paste0("../../data/",herd,"/caribou.gpkg"), "gps", quiet=TRUE)
spring <- y |> filter(migration=="Spring migration")
fall <- y |> filter(migration=="Fall migration")
```

### Spring migration pathways

Each panel shows one individual's spring migration tracks, with years distinguished by color. A few things worth noting from a quick scan:

- Most individuals show a fairly consistent north-to-northeast trajectory across years, though the routes vary.
- A few IDs only have data for one or two years (e.g., some of the later-numbered animals).
- Dark2 has 8 colors, so 6 years are comfortably distinguishable.

```{r}
#| echo: TRUE
#| warning: FALSE
#| message: FALSE
#| fig-width: 10
#| fig-height: 8
spring |>
  st_drop_geometry() |>
  arrange(id, year, timestamp) |>
  ggplot(aes(longitude, latitude, color = factor(year), group = interaction(id, year))) +
  geom_path(linewidth = 0.4) +
  facet_wrap(~id, scales = "free") +
  scale_color_brewer(palette = "Dark2", name = "Year") +
  labs(
    title = "Spring migration",
    x = "Longitude", y = "Latitude"
  ) +
  theme_bw(base_size = 8) +
  theme(
    strip.text = element_text(size = 6),
    axis.text  = element_text(size = 5)
  )
```

```{r}
#| echo: FALSE
#| warning: FALSE
#| message: FALSE
make_season_map(spring, "spring_id_select", "Spring individual:")
```

### Fall migration pathways

Each panel shows one individual's fall migration tracks, with years distinguished by color. A few things worth noting from a quick scan:

- Most individuals show a fairly consistent north-to-northeast trajectory across years, though the routes vary.
- A few IDs only have data for one or two years (e.g., some of the later-numbered animals).
- Dark2 has 8 colors, so 6 years are comfortably distinguishable.

```{r}
#| echo: TRUE
#| warning: FALSE
#| message: FALSE
#| fig-width: 10
#| fig-height: 8
fall |>
  st_drop_geometry() |>
  arrange(id, year, timestamp) |>
  ggplot(aes(longitude, latitude, color = factor(year), group = interaction(id, year))) +
  geom_path(linewidth = 0.4) +
  facet_wrap(~id, scales = "free") +
  scale_color_brewer(palette = "Dark2", name = "Year") +
  labs(
    title = "Fall migration",
    x = "Longitude", y = "Latitude"
  ) +
  theme_bw(base_size = 8) +
  theme(
    strip.text = element_text(size = 6),
    axis.text  = element_text(size = 5)
  )
```

