---
title: Seasonal Migration Pathways
date: 2026-04-29
author: Pierre Vernier
---
  
```{r}
#| echo: TRUE
#| warning: FALSE
#| message: FALSE
library(sf)
library(tidyverse)
library(patchwork)
herd <- "little_rancheria"
```

### Read caribou movement (gps) data

```{r}
#| echo: TRUE
#| warning: FALSE
#| message: FALSE
y <- st_read(paste0("../../data/",herd,"/caribou.gpkg"), "gps", quiet=TRUE)
spring <- y |> filter(migration=="Spring migration")
fall <- y |> filter(migration=="Fall migration")
```

### Spring migration pathways

Each panel shows one individual's spring migration tracks, with years distinguished by color. A few things worth noting from a quick scan:

- Most individuals show a fairly consistent north-to-northeast trajectory across years, though the routes vary.
- A few IDs only have data for one or two years (e.g., some of the later-numbered animals).
- Dark2 has 8 colors, so 6 years are comfortably distinguishable.

```{r}
#| echo: TRUE
#| warning: FALSE
#| message: FALSE
#| fig-width: 10
#| fig-height: 8
spring |>
  st_drop_geometry() |>
  arrange(id, year, timestamp) |>
  ggplot(aes(longitude, latitude, color = factor(year), group = interaction(id, year))) +
  geom_path(linewidth = 0.4) +
  facet_wrap(~id, scales = "free") +
  scale_color_brewer(palette = "Dark2", name = "Year") +
  labs(
    title = "Spring migration",
    x = "Longitude", y = "Latitude"
  ) +
  theme_bw(base_size = 8) +
  theme(
    strip.text = element_text(size = 6),
    axis.text  = element_text(size = 5)
  )
```

### Fall migration pathways

Each panel shows one individual's fall migration tracks, with years distinguished by color. A few things worth noting from a quick scan:

- Most individuals show a fairly consistent north-to-northeast trajectory across years, though the routes vary.
- A few IDs only have data for one or two years (e.g., some of the later-numbered animals).
- Dark2 has 8 colors, so 6 years are comfortably distinguishable.

```{r}
#| echo: TRUE
#| warning: FALSE
#| message: FALSE
#| fig-width: 10
#| fig-height: 8
fall |>
  st_drop_geometry() |>
  arrange(id, year, timestamp) |>
  ggplot(aes(longitude, latitude, color = factor(year), group = interaction(id, year))) +
  geom_path(linewidth = 0.4) +
  facet_wrap(~id, scales = "free") +
  scale_color_brewer(palette = "Dark2", name = "Year") +
  labs(
    title = "Fall migration",
    x = "Longitude", y = "Latitude"
  ) +
  theme_bw(base_size = 8) +
  theme(
    strip.text = element_text(size = 6),
    axis.text  = element_text(size = 5)
  )
```

